---
title: "MoneyBall - NBA data"
author: "Gabriel Majivu"
date: "12/3/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

```{r}
NBA <- read.csv('./data/NBA_train.csv')
str(NBA)
```

Explain the meaning of the variables ...

## Playoffs & Wins

```{r}
table(NBA$W, NBA$Playoffs)
```

From about 42 wins teams always make it to the Playoffs (no zeroes)

```{r}
NBA$PTSdiff <- NBA$PTS - NBA$oppPTS
plot(NBA$PTSdiff, NBA$W)
```

Creating a regression model

```{r}
WinsReg <- lm(W~PTSdiff, data=NBA)
summary(WinsReg)
```

## Points Scored

```{r}
PTSReg <- lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + DRB + TOV + STL + BLK, data=NBA)
summary(PTSReg)
SSE <- sum(PTSReg$residuals^2)
SSE # not so interpretable
RMSE <- sqrt(SSE/nrow(NBA))
RMSE
mean(NBA$PTS)
```

Removing the insignificant variables? TOV - has highest p value

```{r}
PTSReg2 <- lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + DRB + STL + BLK, data=NBA)
summary(PTSReg2)
```

Removing another variable - DRB

```{r}
PTSReg3 <- lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + STL + BLK, data=NBA)
summary(PTSReg3)
```

R-squared remains the same. We are justified in removing the variable.

Removing BLK

```{r}
PTSReg4 <- lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + STL, data=NBA)
summary(PTSReg4)
```

R squared largely remains the same.

```{r}
SSE_4 <- sum(PTSReg4$residuals^2)
SSE_4 # not so interpretable
RMSE_4 <- sqrt(SSE_4/nrow(NBA))
RMSE_4
```

RMSE has not changed much. Seems like we have a much better model after
removing the extra variables.

```{r}
RegVar = NBA[c('X2PA', 'X3PA', 'FTA', 'AST', 'ORB', 'STL')]
cor(RegVar)
```

## Making Predictions

```{r}
NBA_Test <- read.csv('./data/NBA_test.csv')
PointsPrediction <- predict(PTSReg4, newdata=NBA_Test)
SSE <- sum((PointsPrediction - NBA_Test$PTS)^2)
SSE
SST <- sum((mean(NBA$PTS) - NBA_Test$PTS)^2)
SST
R2 <- 1 - SSE/SST
R2
RMSE <- sqrt(SSE/nrow(NBA_Test))
RMSE
```