---
title: "wine_prediction"
author: "Gabriel Majivu"
date: "12/3/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Predicting Wine Quality

Let us see if we can beat the best wine taster at his job!

## Data

```{r}
wine <- read.csv('./data/wine.csv')
str(wine)
summary(wine)
```

## Building a Linear Model

Using one variable

```{r}
model1 <- lm(Price ~ AGST, data=wine)
summary(model1)
```

Errors or residuals:

```{r}
model1$residuals
```

Sum of Squared Errors:

```{r}
SSE <- sum(model1$residuals^2)
SSE
```

Using two variables.

```{r}
model2 <- lm(Price ~ AGST + HarvestRain, data=wine)
summary(model2)
```

R-squared increased from .xx to .7074. Likewise to Adjusted R-sqaured which increased from .xx to 0.6808.
Sum of Squared Errors:

```{r}
SSE <- sum(model2$residuals^2)
SSE
```

Using all variables:

```{r}
model3 <- lm(Price ~ AGST + HarvestRain + WinterRain + Age + FrancePop, data=wine)
summary(model3)
```

R-sqaured value increased to .8294 and Adjusted R-squared to .7845. Adding
these variables improved our model.
Computing Sum of Squared Errors:

```{r}
SSE <- sum(model3$residuals^2)
SSE
```

__Question: Which variables should we keep in the model?__

Fill in some notes here ...

Remove _FrancePop_ variable from the model.

```{r}
model4 <- lm(Price ~ AGST + HarvestRain + WinterRain + Age, data=wine)
summary(model4)
```

R-sqaured decreases slightly, and adjusted R-squared increased. This is a
stronger model than the previous.

Age now has more significance. It is affected by multicollinearity (two 
independent variables are highly correlated) - Age and FrancePop are highly 
correlated.

Computing some correlations

```{r}
cor(wine$Price, wine$WinterRain)
cor(wine$Age, wine$FrancePop)
```

For all variables:

```{r}
cor(wine)
```

Because of multicollinearity, you should remove insignificant variables one at a time.
What woul the model look like if we removed both insignificant variables?

```{r}
model5 <- lm(Price ~ AGST + HarvestRain + WinterRain, data=wine)
summary(model5)
```

R-sqaured and Adjusted R-squared are lower. We would have missed a significant 
variable Age! Age is more easy to interpret in the context of the problem - older wine is
more expensive. So we choose to include it as opposed to have FrancePop variable.
Multicollinearity problem between Age and FrancePop has been fixed. Since there
is no other insignificant variable, we choose to stick with __model4__.

## Making Predictions

We use the test data.

```{r}
wineTest <- read.csv('./data/wine_test.csv')
str(wineTest)
```

Use test data and preferred model to make predictions.

```{r}
predictTest <- predict(model4, newdata=wineTest)
predictTest
```

Computing model accuracy.

```{r}
SSE <- sum((wineTest$Price - predictTest)^2)
SST <- sum((wineTest$Price - mean(wine$Price))^2)
1 - SSE / SST
```

Out of sample accuracy of the model is .7944278. However, the test set is really small.
We should increase it to be more confident.s