---
title: "exploring_kiva_data"
author: "Gabriel Majivu"
date: "11/29/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(stringr)
```

## Data Overview

The data set used in this project is from [kiva](https://www.kiva.org/). The data has these variables:

1. __id__: unique id of the borrower
2. __name__: name of the borrower
3. __status__: current funding status that a loan has. Options include:
    * fundraising - The loan has not yet been funded. You can find the amount funded so far by checking _funded_amount_.
    * funded - This loan request has been completely funded and is not available for new loans by lenders.
    * in_repayment - The loan has been disbursed to the borrowers and they are in the process of using the funds and making payments on the loan to the field partner.
    * paid - The loan has been paid back in full by the borrower.
    * defaulted - A loan which has remained delinquent 6 months after the end of the loan payment schedule.
    * refunded - Refund the funded portion of the loan to lenders after the loan has been partially funded, fully funded, or even during repayment.
4. __funded_amount__: This is the amount of the loan which has been purchased by Kiva lenders.
5. __basket_amount__: This is the amount of the loan which lenders have saved in their shopping baskets, but has not been confirmed as purchased.
6. __activity__: Type of activity the borrower is involved with.
7. __sector__: Type of sector the borrower is involved in.
8. __themes__: General themes that further categorise borrowers.
9. __use__: Description for the purpose of the loan.
10. __partner_id__: Unique id for the field partners.
11. __posted_date__: The date the loan was posted on the Kiva website.
12. __planned_expiration_date__: The date the loan bid expires on the Kiva website. It is not set for some loans.
13. __loan_amount__: Amount of funding sought by the borrowers in a bid for a loan.
14. __borrower_count__: The number of borrowers for a loan. It is a good indication for single vs group borrowers.
15. __lender_count__: Number of lenders who have purchased a loan.
16. __bonus_credit_eligibility__: Whether a loan application is eligible for extra credit.
17. __tags__: Borrower chosen tags to classify their loan applications.
18. __description.languages__: Language choice(s) for the borrower.
19. __image.id__: Image ID for a borrower.
20. __image.template_id__: Image template id.
21. __location.country_code__: Country code of borrower.
22. __location.country__: Country name of borrower.
23. __location.town__ Town of the borrower.
24. __location.geo.level__: Level of accuracy of supplied [geometry](http://www.georss.org/model).
25. __location.geo.pairs__: The coordinate pairs for the geometry.
25. __location.geo.type__: The type of geometry defined by the coordinate pairs provided.
27. __currency_exchange_loss_amount__: Losses realized by the lender due to fluctuations in the value of the local currency against the US dollar. This will result in the paid amount of loan being less than the full loan amount even when the status of the loan is listed as _paid_.
28. __video.id__: Id for video.
29. __video.youtubeId__: Id for video on youtube.
30. __video.title__: Title of the video.
31. __video.thumbnailImageId__: Thumbnail id for the video.

## Loading Data

The data set sourced from [kiva api](http://build.kiva.org/docs/data) contains over 1.1 million loan records.

```{r echo=FALSE}
load('./loans.RData')
```

## Inspecting the loans dataframe

Check the top ten records.

```{r}
head(loans, 10)
```

Check the last 10 records.

```{r}
tail(loans, 10)
```

View the whole dataset.

```{r echo=FALSE}
View(loans)
```

Column names (see Data Overview section for descriptions):

```{r}
names(loans)
```

The shape of the data frame.

```{r}
dim(loans)
```

The structure of the data frame.

```{r}
str(loans)
```

Summary of the dataset

```{r}
summary(loans)
```

Check the number of NA per column

```{r}
character_cols <- names(loans)[sapply(loans, is.character)]
for(col in character_cols){
  na_strings <- sum(loans[[col]]=="NA")
  nas <- sum(is.na(loans[[col]]))
  print(paste(col, ": ", na_strings,',NA:' ,nas))
}
```

```{r}
sapply(loans, FUN = function(x) 100 * sum(is.na(x)) / dim(loans)[1])
```

## Data munging

Remove all columns that we may not need which include: _id_, _image.id_, _image.template_id_, _video.id_, _video.youtubeId_, _video.title_, _video.thumbnailImageId_, _description.languages_

```{r}
# drop variables we do not need
loans$id <- NULL
loans$image.id <- NULL
loans$image.template_id <- NULL
loans$video.id <- NULL
loans$video.youtubeId <- NULL
loans$video.title <- NULL
loans$video.thumbnailImageId <- NULL
#loans$description.languages <- NULL
```

Set loans status as categories

```{r}
loans$status <- factor(loans$status,
                       levels = unique(loans$status),
                       labels = c("Fundraising", "Funded", "Expired"))
```

```{r}
table(loans$status)
```

Set sector as categories

```{r}
loans$sector <- factor(loans$sector)
```

Set latitude and longitude from _location.geo.pairs_

```{r}
loans <- separate(loans, location.geo.pairs, into=c('location.geo.lat', 'location.geo.long'), sep=' ', remove = TRUE)
loans$location.geo.lat <- as.numeric(loans$location.geo.lat)
loans$location.geo.long <- as.numeric(loans$location.geo.long)
```

Add a column to identify direct vs non-direct loans. Direct loans do not require field partners hence partner_id is null.

```{r}
# direct_loans <- loans[is.na(loans$partner_id), ]
# direct_loans$location.country_code
loans$direct <- as.numeric(is.na(loans$partner_id))
```

Add a column to indicate whether loan is borrowed by individual or a group.

```{r}
#loans$group <- as.integer(loans$borrower_count > 1)
```

## Exploratory Plots

### Variable: _status_

```{r}
levels(loans$status)
```

```{r}
df <- data.frame(table(loans$status))
colnames(df) <- c('Status', 'Frequency', 'Percentage')
df$Percentage <- df$Freq / sum(df$Freq) * 100
df
```

The "Expired" status has not been explained on the kiva website. Since it is ambiguous, we shall remove the affected rows from the dataset. Also, there are some loans which have status "Fundraising" which means they [not yet been funded](http://build.kiva.org/docs/data/loans). We will also remove these from the dataset.

```{r}
loans <- subset(loans, status == 'Funded')
```

A plot with the remaining rows:

```{r}
#ggplot(loans, aes(status, ..count..)) + geom_bar()
```

### Variable: _loan_amount_

```{r}
summary(loans$loan_amount)
```

The median is lower than the mean by hundreds of dollars.

```{r}
ggplot(loans, aes(x=loan_amount)) + geom_histogram(binwidth = 200)
```

The plot shows that most of the _loan_amount_ is low. There are high amounts too. To get a better plot, we need to compute a sensible cutoff to remove the high loan amounts.

```{r}
# choosing a number of cutoff amounts
cut_off_amounts <- c(5000, 6000, 7000, 8000, 9000, 10000, 11000, 12000)
for(amount in cut_off_amounts){
    count <- sum(loans$loan_amount > amount)
    proportion <- 100 * count / dim(loans)[1]
    print(paste("cut_off_amount =", amount, ", count =", count,', proportion =' ,proportion, '%'))
}
```

__Notes__:

_loan_amount_ of USD 9,000 is a good pick for large loan amounts cutoff. It only affects 1,598 observations which is about 0.14% of all the records. Further, it is practically a large sum and any amount above it may be safely ignored for purposes of studying financing for the financially underprivileged population.

```{r}
loans <- subset(loans, loans$funded_amount <= 9000)
```

```{r}
ggplot(loans, aes(x=loan_amount)) + geom_histogram(binwidth = 200)
```

Tis is a much clearer plot that shows right skew, with high proportion of lower loan amounts. Most borrowers need small amount capital for their projects.

### Variable: _funded_amount_

```{r}
summary(loans$funded_amount)
```

The lower median and higher mean suggest high value observations that skew the distribution to the right.

```{r}
ggplot(loans, aes(x=funded_amount)) + geom_histogram(binwidth = 200)
```

The plot shows a similar distribution to that of _loan_amount_. The _funded_amount_ is a proportion of the _loan_amount_ bid by the lenders. Perhaps there could be cases where a loan is oversubscribed. We shall see if this is the case in later analysis.

### variable: _borrower_count_

```{r}
summary(loans$borrower_count)
```

```{r}
ggplot(loans, aes(x=borrower_count)) + geom_histogram(binwidth = 3)
```

Most loans have individual borrowers or small groups - few people borrowing together. The _borrower_count_ might be more valuable if it is transformed to categories with specific number of people. Making groups for this variable thus:
* individual  = 1 persons
* small group = 2 - 9 persons
* large group = more than 10 persons

We create a new variable _borrower_group_

```{r}
loans$borrower_group <- cut(loans$borrower_count, c(0,1,9,79), labels=c('Individual','Small','Large'), ordered_result = TRUE)
```

Plotting the new _borrower_group_ variable:

```{r}
ggplot(loans, aes(x=borrower_group)) + geom_bar()
```

How does a box plot of the same variable look like?

```{r}
#ggplot(loans, aes(borrower_group, borrower_count)) + geom_boxplot()
```

The different groups show different variability, the least being the "Individual" and "Large" group showing most variability.
It has some outliers but since the count is small (seen from the preceding plot) it is not necessary to break it down further.

### Variable: _lender_count_

This variable also needs groups. We could apply the same treatment as the borrower count for this group.

```{r}
summary(loans$lender_count)
```

```{r}
ggplot(loans, aes(x=lender_count)) + geom_histogram(binwidth = 5)
```



choose independently.

From the plot we see most loans having fewer than 50 lenders. We can create 3 lender groups thus:
* small group  = less than 10 persons
* medium group = 10 - 50 persons
* large group = more than 50 persons

```{r}
loans$lender_group <- cut(loans$lender_count, c(0,10,50,2986),
                          labels=c('Small', 'medium','Large'),
                          ordered_result = TRUE,
                          include.lowest = TRUE)
```

Plotting the new _borrower_group_ variable:

```{r}
ggplot(loans, aes(x=lender_group)) + geom_bar()
```

THe plot suggests most loans have 10-50 lenders, they attract significant number of lenders. Number of loans with greater than 100 lenders are much fewer than the other two categories combined. Practically high loan amounts are much less and it these that would require many more lenders to combine effort, perhaps minimize their risk.

How does a box plot of the same variable look like?

```{r}
#ggplot(loans, aes(lender_group, borrower_count)) + geom_boxplot()
```

### Variable: _sector_

```{r}
levels(loans$sector)
```

```{r}
table(loans$sector)
```

```{r}
df <- data.frame(table(loans$sector))
df$Percentage <- df$Freq / sum(df$Freq) * 100
colnames(df) <- c('Sector', 'Frequency', 'Percentage')
df
```

From the proportions, Agriculture, Food and Retail sectors are the busiest.

```{r}
ggplot(loans, aes(sector, ..count..)) + geom_bar() + theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0))
```

Most popular sectors include "agriculture", "food", "retail".

### Variable: _bonus_credit_eligibility_

```{r}
levels(loans$bonus_credit_eligibility)
```

```{r}
table(loans$bonus_credit_eligibility)
```

```{r}
ggplot(loans, aes(bonus_credit_eligibility, ..count..)) + geom_bar()
```

### Variable: _funding_rate_

This variable defines the rate of success in funding bid by the borrower, in other words how much of the original sought amount was actually funded?

```{r}
loans$funding_rate <- 100*loans$funded_amount / loans$loan_amount
summary(loans$funding_rate)
```

```{r}
ggplot(loans, aes(x=funding_rate)) + geom_histogram(binwidth=2)
```

As expected, loans with status "funding" have highest _funding_rate_ in comarison with the others which have 0%.

```{r}
sum(loans[["funding_rate"]] < 100)
```

There is only one loan that is underfunded.

```{r}
100*sum(loans[["funding_rate"]] == 100)/dim(loans)[1]
```

99.99% of the loans are 100% funded.
*response variable??*

### Variable: _funding_duration_

This variable defines the period (in days) it takes for a loan to be bought by the lenders.

```{r}
loans$funding_duration <- as.Date(loans$planned_expiration_date) - as.Date(loans$posted_date)
loans$funding_duration <- as.integer(loans$funding_duration)
```

```{r}
summary(loans$funding_duration)
```

The NAs are due to some loan applications missing a _planned_expiration_date_. We could set these to have the median value.

```{r}
mean_duration <- mean(loans$funding_duration, na.rm = TRUE)
loans$funding_duration[is.na(loans$funding_duration)] <- mean_duration
summary(loans$funding_duration)
```

There is an outlier that is skewing this dataseat to the extreme right - it has a very high value for the duration. Ideally such a high duration is not practical since a loan can not stay for that long at a "fundraising stage".

```{r}
sum(loans[["funding_duration"]] > 90)
```


```{r}
ggplot(subset(loans, loans$funding_duration <= 90), aes(x=funding_duration)) + geom_histogram(binwidth = 5)
# transform to a categorical variable - 30 days and more than 30 days
```

The plot shows significant occurences at specific values. We can therefor change this to be a categorical variable with:
* low = less than or equal to 30 days
* high = more than 30 days

```{r}
loans$funding_duration <- cut(loans$funding_duration, c(0, 30, 1674),
                          labels=c('Low','High'),
                          ordered_result = TRUE,
                          include.lowest = TRUE)
```


```{r}
ggplot(loans, aes(funding_duration, ..count..)) + geom_bar()
```

More loans are funded within 30 days.

### Variable: country

Let us try to understand where the applicants originate from.

```{r}
loans$country <- factor(loans$location.country)
table(loans$country)
```

```{r}
df <- data.frame(table(loans$country))
df$Percentage <- df$Freq / sum(df$Freq) * 100
colnames(df) <- c('Country', 'Frequency', 'Percentage')
df
```

Let us plot it for easier visual analysis:

```{r}
ggplot(loans, aes(country, ..count..)) + 
    geom_bar() + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0, lineheight=20, size=6))
```

Kenya and Phillipines have the high counts for borrowers. Perhaps it would be intersting to complement this dataset with one with economic ranking for countries to understand if there is a relationship.

### Variable: description languages

```{r}
loans$num_languages <- str_count(loans$description.languages, ',') + 1
```

```{r}
summary(loans$num_languages)
```

```{r}
ggplot(loans, aes(num_languages, ..count..)) + geom_bar()
```

Most borrower profiles have only one language selected. This is about twice as much as the single language profiles.

### Variable: loan_amount/borrower_count

```{r}
loans$amt_borrower_ratio <- loans$loan_amount / loans$borrower_count
summary(loans$amt_borrower_ratio)
```

```{r}
ggplot(loans, aes(x=amt_borrower_ratio)) + geom_histogram(binwidth = 50)
```

The plot shows a right skew. Most loans have lower _loan_amount_ to _borrower_count_ ration. The bimodal nature oof the plot resembles the distribution for the _loan_amount_ and _funded_amount_ plots seen earlier.

### Variable: loan_amount/lender_count

```{r}
loans$amt_lender_ratio <- loans$loan_amount / loans$lender_count
summary(loans$amt_lender_ratio)
```

```{r}
ggplot(loans, aes(x=amt_lender_ratio)) + geom_histogram(binwidth = 50)
```



---


Q: What is the distribution of funded amounts?

```{r}
ggplot(subset(loans, loans$funded_amount < 10000), aes(x=funded_amount)) + geom_histogram(binwidth = 200)
```

```{r}
# subset
sum(loans$funded_amount > 10000)
```


```{r}
ggplot(subset(loans, loans$funded_amount < 10000), aes(status, funded_amount)) + geom_boxplot()
```

```{r}
ggplot(loans, aes(x=funded_amount, fill=sector, color=sector)) + geom_histogram(binwidth = 1000)
```

__Notes__:
* Right skewed bar plot. Most loans that receive funding are for small amounts. There appears to be a couple of large loan amounts though they must be quite few.
* Do I remove high value loans to aid visualization?
* If yes, what cutoff do I use?

Q Is there any relation between _lender_count_ and _funded_amount_? (Is there a group effect on the lenders with regard to the loan amounts?)

```{r}
ggplot(loans, aes(x=lender_count, y=funded_amount)) + geom_jitter()
```

__Notes__:
There seems to be a relation between number of lenders (_lender_count_) and the loan amounts (_funded_amount_). Higher loan amounts attract more lenders perhaps due to the liability involved and the financial capabilities of the lenders.

Q Are lenders biased against sinlge or goup borrowers?

```{r}
ggplot(loans, aes(x=lender_count, y=borrower_count)) + geom_jitter()

# add a plot between lender count vs single and group
```

__Notes__:
Not sure there is a relationship here.

Q Do groups (multiple borrowers) seek larger amounts as compared to individual borrowers?

new variable -> $/person

```{r}
ggplot(loans, aes(x=funded_amount, y=borrower_count)) + geom_jitter()
```

__Notes__:
Note sure if these are correlated.

Q What is the distribution of the borrowers?

```{r}
ggplot(loans, aes(status, borrower_count)) + geom_boxplot()
```

```{r}
ggplot(loans, aes(sector, borrower_count)) + geom_boxplot() + theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0))
```
visualize on borrower_count and get some direction on it.


Q Does sector the borrower is in affect the loan amounts they attract?

```{r}
ggplot(loans, aes(x=funded_amount, y=sector)) + geom_jitter()
```

Possibly remove high loan amount?
mean funded_amount

__Note__: explore relationship with funded loan amounts and group. Proportion of funded amounts in each sector.