---
title: "Exploring Kiva Loans Data"
author: "Gabriel Majivu"
date: "11/29/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

```{r}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(stringr)
library(scales)
```


### About

[Kiva](https://www.kiva.org/about) is an international non-profit that allows people to lend money to low-income entrepreneurs and students in over 80 countries. Its mission is to enable people to create opportunity for themselves and for others.


### Loans Data Overview

The [data](http://build.kiva.org/docs/data/loans) set used in this project has these variables:

1. __id__: unique id of the borrower
2. __name__: name of the borrower
3. __status__: current funding status that a loan has. Options include:
    * fundraising - The loan has not yet been funded. You can find the amount funded so far by checking _funded_amount_.
    * funded - This loan request has been completely funded and is not available for new loans by lenders.
    * in_repayment - The loan has been disbursed to the borrowers and they are in the process of using the funds and making payments on the loan to the field partner.
    * paid - The loan has been paid back in full by the borrower.
    * defaulted - A loan which has remained delinquent 6 months after the end of the loan payment schedule.
    * refunded - Refund the funded portion of the loan to lenders after the loan has been partially funded, fully funded, or even during repayment.
4. __funded_amount__: This is the amount of the loan which has been purchased by Kiva lenders.
5. __basket_amount__: This is the amount of the loan which lenders have saved in their shopping baskets, but has not been confirmed as purchased.
6. __activity__: Type of activity the borrower is involved with.
7. __sector__: Type of sector the borrower is involved in.
8. __themes__: General themes that further categorise borrowers.
9. __use__: Description for the purpose of the loan.
10. __partner_id__: Unique id for the field partners.
11. __posted_date__: The date the loan was posted on the Kiva website.
12. __planned_expiration_date__: The date the loan bid expires on the Kiva website. It is not set for some loans.
13. __loan_amount__: Amount of funding sought by the borrowers in a bid for a loan.
14. __borrower_count__: The number of borrowers for a loan. It is a good indication for single vs group borrowers.
15. __lender_count__: Number of lenders who have purchased a loan.
16. __bonus_credit_eligibility__: Whether a loan application is eligible for extra credit.
17. __tags__: Borrower chosen tags to classify their loan applications.
18. __description.languages__: Language choice(s) for the borrower.
19. __image.id__: Image ID for a borrower.
20. __image.template_id__: Image template id.
21. __location.country_code__: Country code of borrower.
22. __location.country__: Country name of borrower.
23. __location.town__ Town of the borrower.
24. __location.geo.level__: Level of accuracy of supplied [geometry](http://www.georss.org/model).
25. __location.geo.pairs__: The coordinate pairs for the geometry.
25. __location.geo.type__: The type of geometry defined by the coordinate pairs provided.
27. __currency_exchange_loss_amount__: Losses realized by the lender due to fluctuations in the value of the local currency against the US dollar. This will result in the paid amount of loan being less than the full loan amount even when the status of the loan is listed as _paid_.
28. __video.id__: Id for video.
29. __video.youtubeId__: Id for video on youtube.
30. __video.title__: Title of the video.
31. __video.thumbnailImageId__: Thumbnail id for the video.


### Loading Data

The data was sourced from [kiva api](http://build.kiva.org/docs/data) via [this script](https://github.com/gabeno/data_n00b/blob/master/foundations_of_data_science/capstone_project/data.source.final.R) and saved as an _RData_ file. The data set contains over 1.1 million loan observations.

```{r echo=FALSE}
load('./loans.RData')
```


### Kiva by numbers

A brief look into the data set highlighting important numbers:

```{r}
length(unique(loans$location.country_code))
```

The Kiva dataset has 91 countries represented.

```{r}
length(unique(loans$id))
```

Over 1.1M borrowers have sought funding through Kiva.

```{r}
sum(loans[['funded_amount']])
```

Total amount of funding is $949M.

```{r}
sum(loans[['funded_amount']]) / sum(loans[['loan_amount']]) * 100
```

Repayment rate is 96.4%.


### Inspecting the loans dataframe

Check a sample of the observations

```{r}
head(loans, 10)
```


```{r}
tail(loans, 10)
```

View the whole dataset.

```{r echo=FALSE}
View(loans)
```

Names for columns in the dataset (see Data Overview section for descriptions):

```{r}
names(loans)
```

The dimensions of the loans dataframe.

```{r}
dim(loans)
```

The structure of the loans dataframe.

```{r}
str(loans)
```

Summary of the dataset

```{r}
summary(loans)
```


### Data munging

Check the number of null values (_NAs_) per variable

```{r}
character_cols <- names(loans)[sapply(loans, is.character)]
for(col in character_cols){
  na_strings <- sum(loans[[col]]=="NA")
  nas <- sum(is.na(loans[[col]]))
  print(paste(col, ": ", na_strings,',NA:' ,nas))
}
```

Check the proportion of null values (_NAs_) per variable

```{r}
sapply(loans, FUN = function(x) 100 * sum(is.na(x)) / dim(loans)[1])
```

Remove columns that we may not need which include: _id_, _image.id_, _image.template_id_, _video.id_, _video.youtubeId_, _video.title_, _video.thumbnailImageId_

```{r}
loans$id <- NULL
loans$image.id <- NULL
loans$image.template_id <- NULL
loans$video.id <- NULL
loans$video.youtubeId <- NULL
loans$video.title <- NULL
loans$video.thumbnailImageId <- NULL
```

Each loan has a status at any given time. We then set these loan statuses as factors:

```{r}
loans$status <- factor(loans$status,
                       levels = unique(loans$status),
                       labels = c("Fundraising", "Funded", "Expired"))
table(loans$status)
```

Similarly, we set loan sector as categories:

```{r}
loans$sector <- factor(loans$sector)
table(loans$sector)
```

We also set latitude and longitude from _location.geo.pairs_

```{r}
loans <- separate(loans, location.geo.pairs, into=c('location.geo.lat', 'location.geo.long'), sep=' ', remove = TRUE)
loans$location.geo.lat <- as.numeric(loans$location.geo.lat)
loans$location.geo.long <- as.numeric(loans$location.geo.long)
```

Most loans are administered through Kiva field partners. However, some loans are given directly to the borrowers. These are identified as _direct_ loans. We add a column to identify direct vs non-direct loans. Direct loans do not require field partners hence partner_id is null.

```{r}
loans$direct <- as.numeric(is.na(loans$partner_id))
```


### Exploratory Plots

This section aims to gain insights for some variables of interest.

#### _status_

_status_ informs on the stage a loan has. There are a number of stages as outlined in the Kiva website. Loans for this dataset have the following status:

```{r}
levels(loans$status)
```

Here is the distribution by status of the loan.

```{r}
status_dist <- data.frame(table(loans$status))
status_dist$Percentage <- status_dist$Freq / sum(status_dist$Freq) * 100
colnames(status_dist) <- c('Status', 'Frequency', 'Percentage')
status_dist
```

95.54% of the loans have been funded and only 0.62% are in the fundraising stage. Unfortunately, the "Expired" status has not been explained on the kiva website. Since it is ambiguous, we shall remove the affected rows from the dataset.

```{r}
loans <- subset(loans, status=='Fundraising' | status=='Funded')
```

A plot for the status variable:

```{r}
ggplot(loans, aes(status, ..count..)) +
    geom_bar(width=.2) +
    scale_y_continuous(labels = comma) +
    labs(x = "Loan Status", y = "Count", title = "Loan Status by count") +
    theme(panel.border = element_rect(linetype = "solid", colour = "darkgreen", fill=NA, size = rel(1.2)),
          plot.title = element_text(size = rel(1.5), hjust='0.5', margin = margin(0, 0, 10, 0, 'mm')),
          panel.background = element_rect(fill = "white"),
          panel.grid.major = element_line(colour = "darkgreen", linetype = "solid", size = rel(0.2)),
          panel.grid.minor = element_line(colour = "darkgreen", linetype = "dotted", size = rel(0.2)),
          axis.ticks = element_line(colour = "darkgreen", size = rel(1.2)),
          axis.text.y = element_text(margin = margin(0, 2, 0, 5, 'mm')),
          axis.text.x = element_text(margin = margin(2, 0, 5, 0, 'mm')))
```

#### Variable: _loan_amount_

```{r}
summary(loans$loan_amount)
```

The average amount of loan borrowed is $837.20 We also observe that the median is lower than the mean by a few hundreds of dollars. This points to existence of high loan amounts. The lowest amount requested is $25, the highest being $10,000.

```{r}
ggplot(loans, aes(x=loan_amount)) + geom_histogram(binwidth = 200)
```

As anticipated, there are a number high loan amounts that cause skew on the plot. Since the focus on this study is to evaluate microloans, we need to compute a sensible cutoff to remove the high loan amounts.

```{r}
# choosing a number of cutoff amounts
cut_off_amounts <- c(5000, 6000, 7000, 8000, 9000, 10000, 11000, 12000)
for(amount in cut_off_amounts){
    count <- sum(loans$loan_amount > amount)
    proportion <- 100 * count / dim(loans)[1]
    print(paste("cut_off_amount =", amount, ", count =", count,', proportion =' ,proportion, '%'))
}
```

__Notes__:

For a cutoff at an amount of $9,000 only 1,598 observations will be removed. This represents about 0.14% of data. Further, $9,000 is practically a large sum and any amount above it may be safely ignored for purposes of studying microloans.

```{r}
loans <- subset(loans, loans$loan_amount <= 9000)
summary(loans$loan_amount)
```

With the new dataset, 75% of the loan amounts borrowed are less or equal to $1000 and half the loans are not more than $525. The average amount of loan borrowed is $822.60.

```{r}
ggplot(loans, aes(x=loan_amount)) + geom_histogram(binwidth = 200)
```

Here, it is much clearer that most borrowers go for less than $2,500.

#### Variable: _funded_amount_

```{r}
summary(loans$funded_amount)
```

The average funded amount for a loan is $794.10. We have a naive funding success rate of 96.5% by comparing with the average loan amount. Some loans have $0 funding.

```{r}
ggplot(loans, aes(x=funded_amount)) + geom_histogram(binwidth = 200)
```

The plot shows a similar distribution to that of _loan_amount_. The _funded_amount_ is a proportion of the _loan_amount_ bid by the lenders. Perhaps there could be cases where a loan is oversubscribed. We shall see if this is the case in later analysis.

#### variable: _borrower_count_

```{r}
summary(loans$borrower_count)
```


```{r}
ggplot(loans, aes(x=borrower_count)) + geom_histogram(binwidth = 3)
```

Most loans have individual borrowers or small groups i.e. few people borrowing together. The _borrower_count_ might be more valuable if it is transformed to categories with specific number of people. Making groups for this variable thus:
* individual  = 1 persons
* small group = 2 - 9 persons
* large group = more than 10 persons

We therefore create a new variable _borrower_group_

```{r}
loans$borrower_group <- cut(loans$borrower_count, c(0,1,9,79), labels=c('Individual','Small','Large'), ordered_result = TRUE)
```

Plotting the new _borrower_group_ variable:

```{r}
ggplot(loans, aes(x=borrower_group)) + geom_bar()
```

How does a box plot of the same variable look like?

```{r}
#ggplot(loans, aes(borrower_group, borrower_count)) + geom_boxplot()
```

The different groups show different variability, the least being the "Individual" and "Large" group showing most variability.
It has some outliers but since the count is small (seen from the preceding plot) it is not necessary to break it down further.

#### Variable: _lender_count_

This variable also needs groups. We could apply the same treatment as the borrower count for this group.

```{r}
summary(loans$lender_count)
```

Lender group average 23 persons per loan. Lenders tend to spread their risk by grouping and buying into loans in small amounts

```{r}
ggplot(loans, aes(x=lender_count)) + geom_histogram(binwidth = 5)
```



choose independently.

From the plot we see most loans having fewer than 50 lenders. We can create 3 lender groups thus:
* small group  = less than 10 persons
* medium group = 10 - 50 persons
* large group = more than 50 persons

```{r}
loans$lender_group <- cut(loans$lender_count, c(0,10,50,2986),
                          labels=c('Small', 'medium','Large'),
                          ordered_result = TRUE,
                          include.lowest = TRUE)
```

Plotting the new _borrower_group_ variable:

```{r}
ggplot(loans, aes(x=lender_group)) + geom_bar()
```

THe plot suggests most loans have 10-50 lenders, they attract significant number of lenders. Number of loans with greater than 100 lenders are much fewer than the other two categories combined. Practically high loan amounts are much less and it these that would require many more lenders to combine effort, perhaps minimize their risk.

#### Variable: _sector_

```{r}
levels(loans$sector)
```

```{r}
table(loans$sector)
```

```{r}
sector_dist <- data.frame(sort(table(loans$sector), decreasing = TRUE))
sector_dist$Percentage <- sector_dist$Freq / sum(sector_dist$Freq) * 100
colnames(sector_dist) <- c('Sector', 'Frequency', 'Percentage')
sector_dist
```

```{r}
ggplot(sector_dist, aes(Sector, Frequency)) + 
    geom_bar(stat="identity") + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0))
```

Most popular sectors include "agriculture", "food", "retail". Providing food for the family is an important thing for most breadwinners and this conforms to what the data shows. Retail also is popular because of small businesses that people have.

#### Variable: _bonus_credit_eligibility_

```{r}
levels(loans$bonus_credit_eligibility)
```

```{r}
table(loans$bonus_credit_eligibility)
```

```{r}
ggplot(loans, aes(bonus_credit_eligibility, ..count..)) + geom_bar()
```

#### Variable: _funding_rate_

This variable defines the rate of success in funding bid by the borrower, in other words how much of the original sought amount was actually funded?

```{r}
loans$funding_rate <- 100*loans$funded_amount / loans$loan_amount
summary(loans$funding_rate)
```

The average rate of funding is 97.69% with the highest being 113.30%. Funding rate is considered as the proportion of loan which a lender bought. Any value more than 100% was overfunded.

```{r}
ggplot(loans, aes(x=funding_rate)) + geom_histogram(binwidth=2)
```

```{r}
sum(loans[["funding_rate"]] <= 100)
```

About 1.1M loans are underfunded and ...

```{r}
sum(loans[["funding_rate"]] > 100)
```

only two are overfunded.

### Variable: _funding_duration_

This variable defines the period (in days) it takes for a loan to be bought by the lenders.

```{r}
loans$funding_duration <- as.Date(loans$planned_expiration_date) - as.Date(loans$posted_date)
loans$funding_duration <- as.integer(loans$funding_duration)
```

```{r}
summary(loans$funding_duration)
```

The NAs are due to some loan applications missing a _planned_expiration_date_. We could set these to have the median value.

```{r}
mean_duration <- mean(loans$funding_duration, na.rm = TRUE)
loans$funding_duration[is.na(loans$funding_duration)] <- mean_duration
summary(loans$funding_duration)
```

There is a loan which has a very long duration. Ideally such a high duration is not practical since a loan can not stay for that long at a "fundraising stage". If we consider a maximum period of 90 days as the funding period, then 339 observations are affected. 90 days is reasonable because it is a common credit period in the finance world.

```{r}
sum(loans[["funding_duration"]] > 90)
```


```{r}
ggplot(subset(loans, loans$funding_duration <= 90), aes(x=funding_duration)) + geom_histogram(binwidth = 5)
# transform to a categorical variable - 30 days and more than 30 days
```

The plot shows significant occurences at specific values. We can therefor change this to be a categorical variable with:
* low = less than or equal to 30 days
* high = more than 30 days

```{r}
loans$funding_duration <- cut(loans$funding_duration, c(0, 30, 1674),
                          labels=c('Low','High'),
                          ordered_result = TRUE,
                          include.lowest = TRUE)
```


```{r}
ggplot(loans, aes(funding_duration, ..count..)) + geom_bar()
```

More loans are funded within 30 days. Perhaps this goes with the thought that most loan amounts are small thereby they do not need a lot of "funding time".

#### Variable: country

Let us try to understand where the applicants originate from.

```{r}
loans$country <- factor(loans$location.country)
table(loans$country)
```

```{r}
country_dist <- data.frame(sort(table(loans$country), decreasing = TRUE))
country_dist$Percentage <- country_dist$Freq / sum(country_dist$Freq) * 100
colnames(country_dist) <- c('Country', 'Frequency', 'Percentage')
country_dist
```

Let us plot it for easier visual analysis:

```{r, fig.width = 8, fig.height = 6}
ggplot(country_dist, aes(Country, Frequency)) + 
    geom_bar(stat="identity") + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0, size=12))
```

Kenya and Phillipines have the high counts for borrowers. Perhaps it would be intersting to complement this dataset with one with economic ranking for countries to understand if there is a relationship.

The top ten countries by loan count:
 - Phillipines (19.4%)
 - Kenya (10.0%)
 - Peru (6.78%)
 - Cambodia (5.34%)
 - El Salvador (4.53%)
 - Uganda (3.34%)
 - Nicaragua (3.28%)
 - Tajikistan (2.98%)
 - Pakistan (2.86%)
 - Equador (2.38%)
 
```{r}
country_top_ten <- country_dist[1:10,]
ggplot(country_top_ten, aes(Country, Frequency), fill=loans$sector) + 
    geom_bar(stat="identity") + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0, size=12))
```
 

```{r, fig.width = 8, fig.height = 6}
loans_top_ten <- subset(loans, loans$country %in% country_top_ten[['Country']])
ggplot(loans_top_ten, aes(country, ..count.., fill=sector)) + 
    geom_bar(position='dodge') + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0, size=12))
```

### Variable: description languages

```{r}
loans$num_languages <- str_count(loans$description.languages, ',') + 1
```

```{r}
summary(loans$num_languages)
```

```{r}
ggplot(loans, aes(num_languages, ..count..)) + geom_bar()
```

Most borrower profiles have only one language selected. This is about twice as much as the single language profiles.

### Variable: loan_amount/borrower_count

```{r}
loans$amt_borrower_ratio <- loans$loan_amount / loans$borrower_count
summary(loans$amt_borrower_ratio)
```

```{r}
ggplot(loans, aes(x=amt_borrower_ratio)) + geom_histogram(binwidth = 50)
```

The plot shows a right skew. Most loans have lower _loan_amount_ to _borrower_count_ ration. The bimodal nature oof the plot resembles the distribution for the _loan_amount_ and _funded_amount_ plots seen earlier.

### Variable: loan_amount/lender_count

```{r}
loans$amt_lender_ratio <- loans$loan_amount / loans$lender_count
summary(loans$amt_lender_ratio)
```

```{r}
ggplot(loans, aes(x=amt_lender_ratio)) + geom_histogram(binwidth = 50)
```

## Explore Relationships

```{r}
ggplot(loans, aes(x=loan_amount, y=funded_amount)) + geom_jitter()
```

_funded_amount_ is a proportion of the _loan_amount_ so they are expected to have a strong correlation. The plot above display this.

```{r}
ggplot(loans, aes(x=borrower_count, y=funded_amount)) + geom_jitter()
```

Not much to deduce from this plot.

```{r}
ggplot(loans, aes(x=lender_count, y=funded_amount)) + geom_jitter()
```

Higher loan amounts attract more lenders. It is an indication that most lenders buy small chunks of a loan.

```{r}
ggplot(loans, aes(x=lender_count, y=borrower_count)) + geom_jitter()
```

Nothing to deduce in this plot.

```{r, fig.width = 8, fig.height = 6}
ggplot(loans, aes(x=sector, y=funded_amount)) + 
    geom_boxplot() + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0, size=12))
```

75% of borrowing in all sectors is less than $1,250.

```{r}
ggplot(loans, aes(sector, ..count.., color=lender_group, fill=lender_group)) + 
    geom_bar(position="dodge") + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0))
```

"medium" _lender_group_ dominates the sectors.

```{r}
ggplot(loans, aes(sector, ..count.., color=borrower_group, fill=borrower_group)) + 
    geom_bar(position="dodge") + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0))
```

Indivudual borrowers are dominant across all sectors.

```{r, fig.width = 8, fig.height = 6}
ggplot(loans, aes(x=country, y=mean(funded_amount))) + 
    geom_col() + 
    theme(axis.text.x = element_text(angle=90, hjust=0, vjust=0, size=12))
```

Countries with high number of borrowing - Cambodia, Kenya and Phillipines - have low variablity in the amounts loaned. Most people borrow small amounts of money. They are developing economies and imply low cadre pf borrowers trying to raise funds for their projects. 75% of the loans borrowed in these countries is less than $1,250. United States has a high variability and the median loan amount is $5,000. In other countries though, the borrowing is quite varied, e.g Congo and Myanmar.